Hierarchy Builder: algebraic~hierarchies made easy in Coq with Elpi

### 2020-03-11 hb.tex
fix minor issues

In this paper we describe \HB{}, a high level language
to \emph{build} hierarchies of algebraic structures and to make these hierarchies
\emph{evolve} without breaking user code. The key concepts are the ones of
\emph{\factory{}} and \emph{\phantterm{}} that let the hierarchy developer describe an actual
interface for their library. Behind that interface the developer can provide
appropriate code to ensure backward compatibility.
We implement the \HB{} language by compiling it
to a variant of the Packed Classes discipline, that we call \emph{flat},
in the \hb{} addon for Coq. We write this addon using the
Elpi~\cite{DBLP:conf/lpar/DunchevGCT15,CoqElpi} extension language.\\

@@

In order to build a structure we need to declare some \factories{} and
later assemble them. One kind of \factory{} supported by \HB{}, the simplest
one, is called \emph{\mixin{}} and is embodied by a record, by convention named
\coq{axioms}, that gathers operations and properties.

\Mixins{} are declared via the \hbmixin{} command
that takes a record declaration with
a type variable name and a possibly empty list of \factories{} for that type.
The code between lines~\ref{demo:mixin:mot}
and~\ref{demo:mixin:mot:end} declares a \mixin{} that can turn a naked
type \coq{A} into a monoid, hence we chose the name to be \coq{Monoid_of_Type}.

The \hbstructure{} command takes in input the definition S for a type
\coq{A} together with a list of \factories{} for \coq{A}.
It registers a structure S in the hierarchy placing
any definition specific to that structure inside a Coq module named \coq{S}.
Line~\ref{demo:structure:monoid} hence forges the \coq{Monoid} structure.

@@

We use the \coq{Monoid_of_Type.Build} \emph{\phantterm{}}
(line~\ref{demo:theory:z:monoid:axioms})
in order to build an instance of the \coq{Monoid}
structure for binary integers \coq{Z}.
We then register that monoid instance as the canonical one on \coq{Z}
(line~\ref{demo:theory:z:monoid:canonical}) via the command
\coq{hb.canonical}.
We can similarly declare that \coq{Z} forms a ring by using
the \coq{Ring_of_Monoid.Build} \phantterm{}
(lines~\ref{demo:theory:z:ring:axioms} and~\ref{demo:theory:z:ring:canonical}).
Note that the \coq{Ring_of_Monoid.Build} \phantterm{} is not
a plain record constructor for \coq{Ring_of_Monoid}, since that
would require more arguments, namely the monoid ones (see the \coq{_}
at line~\ref{demo:theory:z:ring:axioms}). The
\phantterm{} synthesized by \HB{} infers them automatically
(as in~\cite[Section 7]{DBLP:conf/itp/MahboubiT13}) thanks to the \coq{hb.canonical}
declaration given just above.

>>>>

In this paper we describe \HB{}, a high level language
to \emph{build} hierarchies of algebraic structures and to make these hierarchies
\emph{evolve} without breaking user code. The key concepts are the ones of
\emph{\factory{}}, \emph{\builder{}} and \emph{\phantterm{}} that let the hierarchy developer describe an actual
interface for their library. Behind that interface the developer can provide
appropriate code to ensure backward compatibility.
We implement the \HB{} language by compiling it
to a variant of the Packed Classes discipline, that we call \emph{flat},
in the \hb{} addon for Coq. We write this addon using the
Elpi~\cite{DBLP:conf/lpar/DunchevGCT15,CoqElpi} extension language.\\

@@

In order to build a structure we need to declare some \factories{} and
later assemble them. One kind of \factory{} supported by \HB{}, the simplest
one, is called \emph{\mixin{}} and is embodied by a record
that gathers operations and properties.

\Mixins{} are declared via the \hbmixin{} command
that takes a record declaration with
a type parameter and a possibly empty list of \factories{} for that type.
The code between lines~\ref{demo:mixin:mot}
and~\ref{demo:mixin:mot:end} declares a \mixin{} that can turn a naked
type \coq{A} into a monoid, hence we chose the name to be \coq{Monoid_of_Type}.

The \hbstructure{} command takes in input a definition for a sigma type S
that equips \coq{A} with a list of \factories{}.
It registers a structure S in the hierarchy placing
any definition specific to that structure inside a Coq module named \coq{S}.
Line~\ref{demo:structure:monoid} hence forges the \coq{Monoid} structure.

@@

We use the \coq{Monoid_of_Type.Build} \emph{\phantterm{}}
(line~\ref{demo:theory:z:monoid:axioms})
in order to build an instance of the \coq{Monoid}
structure for binary integers \coq{Z}.
We then register that monoid instance as the canonical one on \coq{Z}
(line~\ref{demo:theory:z:monoid:canonical}) via the command
\coq{hb.canonical}.
We can similarly declare that \coq{Z} forms a ring by using
the \coq{Ring_of_Monoid.Build} \phantterm{}
(lines~\ref{demo:theory:z:ring:axioms} and~\ref{demo:theory:z:ring:canonical}).
Note that the \coq{Ring_of_Monoid.Build} \phantterm{} is not
a plain record constructor for \coq{Ring_of_Monoid}, since that
would require more arguments, namely the monoid ones (see the \coq{_}
at line~\ref{demo:theory:z:ring:axioms}). The
\phantterm{} synthesized by \HB{} infers them automatically
(as in~\cite[Section 7]{DBLP:conf/itp/MahboubiT13}) thanks to the \hbinstance{}
declaration given just above.
